Predicting tree failure likelihood for utility risk mitigation via artificial intelligence

### 2021-03-05 docs/manuscript/manuscript.tex
Updates from Overleaf

\begin{abstract}
  Critical to the resilience of utility power lines, tree failure assessments have historically been performed via
  manual and visual inspections.  In this paper, we develop a convolutional neural network (CNN) to predict tree failure
  likelihood categories (\textit{Probable}, \textit{Possible}, \textit{Improbable}) under four classification scenarios.
  Starting with an original set of 505 expert-labeled images, we perform preprocessing and augmentation tasks to
  increase the number of samples. We optimize several hyperparameters in our CNN and then train and test its performance
  for three different image resolutions. The trained CNN produced a validation accuracy of at least 0.94
  ($\hat\sigma = 0.1)$ in the best-performing, yet hypothetical scenario as it excludes one of the categories.  The
  second-best performing scenario, which includes all categories and therefore more practical, resulted in a validation
  accuracy of 0.92 ($\hat\sigma = 0.1$). Thus, via this novel framework, we demonstrate the potential of artificial
  intelligence to automate and consequently reduce the costs of tree failure likelihood assessments, thereby promoting
  sustainable infrastructure.
\end{abstract}

@@

The various hyperparameters in the model are summarized in \autoref{tab:hyp}.
\begin{table}[]\small
\rowcolors{1}{white}{gray!35}
    \centering
    \begin{tabular}{l c c c c c c c}\toprule
    \bf Layer & \bf Layer \# & \bf No.\ Filters & \bf Kernel Size & \bf Strides &
    \bf Activation &   \bf Rate & \bf No.\ Units  \\\midrule
    Convolutional  & 1         & 64                & $k^*$   & 2           & ReLU && \\
    Max. Pooling   & 1         &                   & 2               &&&             &      \\
    Convolutional  & 2         & 128               & 3   & 1           & ReLU && \\
    Convolutional  & 3         & 128               & 3   & 1           & ReLU && \\
    Max. Pooling   & 2         &                   & 2               &&&             &      \\    
    Convolutional  & 4         & 256               & 3   & 1           & ReLU && \\
    Convolutional  & 5         & 256               & 3   & 1           & ReLU && \\
    Max. Pooling   & 3         &                   & 2               &             &&&      \\\midrule
    Flatten & & & & &&& \\\midrule
%     &  &   &  & & &
%     \bf Rate & \bf No.\ Units  \\\midrule
    Dense           & 1  & & &  &  $a_1^*$ & & $u_1^*$ \\
    Dropout         & 1  & & &  && $r_1^*$ &   \\
    Dense           & 2  & & &  &  $a_2^*$ & & $u_2^*$  \\
    Dropout         & 2  & & &  && $r_2^*$ &  \\ \bottomrule
    \end{tabular}
    \caption{Summary of the convolutional neural network hyperparameters. 
    Those indicated by an asterisked symbol are optimized using a guided search.}
    \label{tab:hyp}
\end{table}

@@

\subsection{Sensitivity to training resolution}
\autoref{fig:sens-val-acc} shows the boxplots of the validation performance metrics for all twelve cases considered.
The metrics are: accuracy and the macro-averages of precision, recall and the $F_{1}$ score.
%With respect to the accuracy, there are no significant differences between the 
Taking all metrics into consideration, Welch pairwise tests showed that there are no significant differences in performance among the 3 training resolutions, with one exception.
In the scenario Pr\_PoIm, the accuracy, $F_{1}^{m}$, $Re^{m}$ for the 128-pixel case are greater than those for the 224-pixel case ($p<.001$).
This difference in performance is not as stark between the 128-pixel and 64-pixel cases in the same scenario.
This outcome implies that we can achieve efficiency by  training at lower resolutions without significant losses in performance.

@@

\subsection{Analysis of classification strategies}
Given that there was no significant difference in performance among the three resolutions tested, we focused our
attention on the 128-px case, as a tradeoff between efficiency and performance. We re-trained the model on all the four
scenarios using this resolution and then evaluated performance on the afore-mentioned validation metrics: accuracy and the macro-averages of
class-specific precision, recall and $F_{1}$ score.
The goal of this analysis was to explore the efficiencies of various classification strategies for tree failure risk.

Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the
respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios. The
metrics are shown in \autoref{fig:cnn-perf}.  In each scenario, we trained the model over 13 epochs. % but we employed
% early stopping using validation loss as the criterion with a patience of 3.
% That is, the training routine would set the
% weights of the final model at the epoch from which there was no further improvement in validation loss after three
% subsequent iterations.
From this figure, we see that Pr\_Po\_Im performs significantly worse than the other three. This indicates the
uncertainty surrounding the expert, yet subjective, failure likelihood assessment of the trees to begin
with, particularly when the category is deemed to be \textit{Possible}. % Given the results
% from the sensitivity tests, however, it is possible that a different training instance may have provided better
% results, or perhaps, further iterations were required to achieve convergence.

@@

We also plotted the confusion matrix for each scenario in \autoref{fig:confusion-matrix} to further investigate the
performance of the classifier.  Each row of the confusion matrix indicates the proportion of observations in a given
class that are predicted to be in the classes across the columns.  Thus, the diagonal entry in each matrix represents
the class-specific recall score, $Re_{c}$.  The matrices show that generally, the \textit{Possible} category is
difficult to predict accurately as an individual class.  The classifier performs best when it only has to distinguish
between \textit{Probable} and \textit{Improbable} (\autoref{pr_im_cm}). Performance only suffers slightly when
\textit{Possible} was grouped with \textit{Probable} in the PrPo\_Im scenario (\autoref{prpo_im_cm}), which we would
select as the best strategy from a practical standpoint.  These two scenarios (Pr\_Im and PrPo\_Im) have the highest
class-specific recall scores $Re_{c}$ ($\ge 0.90$) compared to the other two scenarios Pr\_PoIm ($Re_{c} \ge 0.58$) and
Pr\_Po\_Im ($Re_{c} \ge 0.54)$. In scenario Pr\_PoIm, we see that even grouping \textit{Possible} with
\textit{Improbable} worsens the ability of the classifier to distinguish the \textit{Probable} category.

@@

\section{Acknowledgments}
The authors acknowledge the partial support of Eversource in funding this work.

>>>>

\begin{abstract}
  Critical to the resilience of utility power lines, tree failure assessments have historically been performed via
  manual and visual inspections.  In this paper, we develop a convolutional neural network (CNN) to predict tree failure
  likelihood categories (\textit{Probable}, \textit{Possible}, \textit{Improbable}) under four classification scenarios.
  Starting with an original set of 505 expert-labeled images, we performed preprocessing and augmentation tasks to
  increase the number of samples by a factor of 5. We optimized several hyperparameters in the CNN and then trained and assessed its performance via cross-validation
  for three different image resolutions under all four scenarios. The CNN produced a validation accuracy of at least 0.94
  ($\hat\sigma = 0.1)$ in the best-performing, yet hypothetical scenario as it excludes one of the categories.  The
  second-best performing scenario, which includes all categories and therefore more practical, resulted in a validation
  accuracy of 0.92 ($\hat\sigma = 0.1$). Thus, via this novel framework, we demonstrate the potential of artificial
  intelligence to automate and consequently reduce the costs of tree failure likelihood assessments, thereby promoting
  sustainable infrastructure.
\end{abstract}

@@

The various hyperparameters in the model are summarized in \autoref{tab:hyp}.
\begin{table}[]\small
\rowcolors{1}{white}{gray!35}
    \centering
    \begin{tabular}{l c c c c c c c}\toprule
    \bf Layer & \bf Layer \# & \bf No.\ Filters & \bf Kernel Size & \bf Strides &
    \bf Activation &   \bf Rate & \bf No.\ Units  \\\midrule
    Convolutional  & 1         & 64                & $k$   & 2           & ReLU && \\
    Max. Pooling   & 1         &                   & 2               &&&             &      \\
    Convolutional  & 2         & 128               & 3   & 1           & ReLU && \\
    Convolutional  & 3         & 128               & 3   & 1           & ReLU && \\
    Max. Pooling   & 2         &                   & 2               &&&             &      \\    
    Convolutional  & 4         & 256               & 3   & 1           & ReLU && \\
    Convolutional  & 5         & 256               & 3   & 1           & ReLU && \\
    Max. Pooling   & 3         &                   & 2               &             &&&      \\\midrule
    Flatten & & & & &&& \\\midrule
%     &  &   &  & & &
%     \bf Rate & \bf No.\ Units  \\\midrule
    Dense           & 1  & & &  &  $a_1$ & & $u_1$ \\
    Dropout         & 1  & & &  && $r_1$ &   \\
    Dense           & 2  & & &  &  $a_2$ & & $u_2$  \\
    Dropout         & 2  & & &  && $r_2$ &  \\ \bottomrule
    \end{tabular}
    \caption{Summary of the convolutional neural network hyperparameters. 
    Those represented by a symbol are optimized using a guided search.}
    \label{tab:hyp}
\end{table}

@@

\subsection{Sensitivity to training resolution}
\autoref{fig:sens-val-acc} shows the boxplots of the validation performance metrics for all twelve cases considered.
The metrics are: accuracy and the macro-averages of precision, recall and the $F_{1}$ score.
%With respect to the accuracy, there are no significant differences between the 
Taking all metrics into consideration, Welch pairwise tests showed that there are no significant differences in performance among the 3 training resolutions, with one exception.
In the scenario Pr\_PoIm, the accuracy, $F_{1}^{m}$, and $Re^{m}$ for the 128-pixel case are greater than those for the 224-pixel case ($p<.001$).
This difference in performance is not as stark between the 128-pixel and 64-pixel cases in the same scenario.
This outcome implies that we can achieve efficiency by  training at lower resolutions without significant losses in performance.

@@

\subsection{Analysis of classification strategies}
Given that there was no significant difference in performance among the three resolutions tested, we focused 
on the 128-px case, as a tradeoff between efficiency and performance. We re-trained the model on all four
scenarios using this resolution and then evaluated performance on the aforementioned validation metrics: accuracy and the macro-averages of
class-specific precision, recall and $F_{1}$ score.
The goal of this analysis was to explore the efficiencies of various classification strategies for tree failure risk.

Using a randomly sampled validation set that was 20\% of the augmented dataset as before, we trained the CNN with the
respective optimal hyperparameters for a single instance in order to compare the performance across the scenarios. The
metrics are shown in \autoref{fig:cnn-perf}.  In each scenario, we trained the model over 13 epochs. % but we employed
% early stopping using validation loss as the criterion with a patience of 3.
% That is, the training routine would set the
% weights of the final model at the epoch from which there was no further improvement in validation loss after three
% subsequent iterations.
From this figure, we see that Pr\_Po\_Im performed significantly worse than the other three. This indicates the
uncertainty surrounding the expert, yet subjective, failure likelihood assessment of the trees to begin
with, particularly when the category is deemed to be \textit{Possible}. % Given the results
% from the sensitivity tests, however, it is possible that a different training instance may have provided better
% results, or perhaps, further iterations were required to achieve convergence.

@@

We also plotted the confusion matrix for each scenario in \autoref{fig:confusion-matrix} to further investigate the
performance of the classifier.  Each row of the confusion matrix indicates the proportion of observations in a given
class that are predicted to be in the classes across the columns.  Thus, the diagonal entry in each matrix represents
the class-specific recall score, $Re_{c}$.  The matrices show that generally, the \textit{Possible} category is
difficult to predict accurately as an individual class.  The classifier performs best when it only has to distinguish
between \textit{Probable} and \textit{Improbable} (\autoref{pr_im_cm}). This observation aligns with previous work \cite{koeser2020can} and is arboriculturally intuitive. The \textit{Improbable} category includes trees with no apparent structural defects, which is comparatively easy for a qualified arborist to assess. When a defect is present, however, an arborist must estimate the severity of the defect that is present, whether the tree has grown in such a way that structurally compensates for the presence of the defect, and the anticipated loads on the tree. Although there are guidelines to inform an arborist's assessment \cite{smiley2017best,goodfellow2020best}, evaluating each of these factors still depends on an arborist's judgment and experience, making it more difficult to distinguish between trees in the \textit{Possible} and \textit{Probable} categories. Thus, it was not surprising that performance only suffered slightly when
\textit{Possible} was grouped with \textit{Probable} in the PrPo\_Im scenario (\autoref{prpo_im_cm}), which we would
select as the best strategy from a practical standpoint.  These two scenarios (Pr\_Im and PrPo\_Im) have the highest
class-specific recall scores $Re_{c}$ ($\ge 0.90$) compared to the other two scenarios Pr\_PoIm ($Re_{c} \ge 0.58$) and
Pr\_Po\_Im ($Re_{c} \ge 0.54)$. In scenario Pr\_PoIm, we see that even grouping \textit{Possible} with
\textit{Improbable} worsens the ability of the classifier to distinguish the \textit{Probable} category.

@@

\section{Acknowledgments}
The authors acknowledge the partial support of Eversource Energy in funding this work.
